---
title: "eda"
output: html_document
date: "2026-04-08"
---

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
```

# 2. Read the dataset

```{r}
crop_data <-read_csv("C:/Users/T00759558/Downloads/Crop_Recommendation.csv")
```

# 3. View the basic structure of the data

```{r}
str(crop_data)
head(crop_data)
dim(crop_data)
names(crop_data)

```

# 4. Check missing values



```{r}
colSums(is.na(crop_data))
```

# 5. Check duplicated rows

```{r}
sum(duplicated(crop_data))
```

# 6. Convert Crop into factor


```{r}
crop_data$Crop <- as.factor(crop_data$Crop)
```


# 7. Summary statistics

```{r}
summary(crop_data)
```

# 8. Count the number of crop classes

```{r}
table(crop_data$Crop)
```
# 9. Plot crop class distribution

```{r}
ggplot(crop_data, aes(x = Crop)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Crop Types",
    x = "Crop Type",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# 10. Histograms for numeric variables

```{r}
ggplot(crop_data, aes(x = Nitrogen)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Nitrogen")

ggplot(crop_data, aes(x = Phosphorus)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Phosphorus")

ggplot(crop_data, aes(x = Potassium)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Potassium")

ggplot(crop_data, aes(x = Temperature)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Temperature")

ggplot(crop_data, aes(x = Humidity)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Humidity")

ggplot(crop_data, aes(x = pH_Value)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of pH Value")

ggplot(crop_data, aes(x = Rainfall)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Rainfall")

```



# 11. Boxplots to check spread and possible outliers


```{r}
ggplot(crop_data, aes(y = Nitrogen)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Nitrogen")

ggplot(crop_data, aes(y = Phosphorus)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Phosphorus")

ggplot(crop_data, aes(y = Potassium)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Potassium")

ggplot(crop_data, aes(y = Temperature)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Temperature")

ggplot(crop_data, aes(y = Humidity)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Humidity")

ggplot(crop_data, aes(y = pH_Value)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of pH Value")

ggplot(crop_data, aes(y = Rainfall)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Rainfall")

```

# 12. Correlation matrix for numeric variables


```{r}
numeric_data <- crop_data %>%
  dplyr::select(Nitrogen, Phosphorus, Potassium, Temperature, Humidity, pH_Value, Rainfall)

cor_matrix <- cor(numeric_data)
print(cor_matrix)
```

# 13. Correlation heatmap


```{r}
cor_df <- as.data.frame(as.table(cor_matrix))

ggplot(cor_df, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Freq, 2)), size = 4) +
  labs(
    title = "Correlation Heatmap of Numeric Variables",
    x = "",
    y = "",
    fill = "Correlation"
  ) +
  theme_minimal()

```

# 14. Scatterplots for some important relationships

```{r}
ggplot(crop_data, aes(x = Nitrogen, y = Phosphorus, color = Crop)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Nitrogen vs Phosphorus by Crop",
    x = "Nitrogen",
    y = "Phosphorus"
  ) +
  theme_minimal()

ggplot(crop_data, aes(x = Temperature, y = Humidity, color = Crop)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Temperature vs Humidity by Crop",
    x = "Temperature",
    y = "Humidity"
  ) +
  theme_minimal()

ggplot(crop_data, aes(x = Rainfall, y = pH_Value, color = Crop)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Rainfall vs pH Value by Crop",
    x = "Rainfall",
    y = "pH Value"
  ) +
  theme_minimal()
```

# 15. Mean values of numeric variables by crop

```{r}
crop_means <- crop_data %>%
  group_by(Crop) %>%
  summarise(
    Mean_Nitrogen = mean(Nitrogen),
    Mean_Phosphorus = mean(Phosphorus),
    Mean_Potassium = mean(Potassium),
    Mean_Temperature = mean(Temperature),
    Mean_Humidity = mean(Humidity),
    Mean_pH = mean(pH_Value),
    Mean_Rainfall = mean(Rainfall)
  )

print(crop_means)

```

# 16. Prepare long-format data for grouped mean plot

```{r}
crop_means_long <- crop_means %>%
  pivot_longer(
    cols = -Crop,
    names_to = "Variable",
    values_to = "Mean_Value"
  )
```

# 17. Plot average values of variables by crop


```{r}
ggplot(crop_means_long, aes(x = Crop, y = Mean_Value, fill = Variable)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Variable Values by Crop",
    x = "Crop",
    y = "Mean Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
