---
title: "Project"
date: "2026-04-06"
output: html_document
---


```{r}
library(dplyr)
library(ggplot2)
library(caret)
library(rsample)
library(rpart)
library(rpart.plot)
library(randomForest)
library(ranger)
library(cluster)
library(class)
```

# 1. Read the crop dataset

```{r}
crop_data <- read.csv("Crop_Recommendation.csv")

# Check the structure
str(crop_data)
head(crop_data)

# Make sure the response variable is a factor
crop_data$Crop <- as.factor(crop_data$Crop)
```

# 2.Keep only the variables needed

```{r}
crop_data <- crop_data %>%
  dplyr::select(
    Nitrogen,
    Phosphorus,
    Potassium,
    Temperature,
    Humidity,
    pH_Value,
    Rainfall,
    Crop
  )
```

# 3. Group the observations using K-means

```{r}
# K-means uses only numeric predictors, not the crop label
cluster_data <- crop_data %>%
  dplyr::select(
    Nitrogen,
    Phosphorus,
    Potassium,
    Temperature,
    Humidity,
    pH_Value,
    Rainfall
  )

# Standardize the variables before clustering
cluster_data_scaled <- scale(cluster_data)
```

# 4. Create 4 clusters

```{r}
set.seed(4050)

best_k <- 4
best_k
```
# 5.Fit the final K-means model


```{r}
set.seed(4050)
km_final <- kmeans(cluster_data_scaled, centers = best_k, nstart = 25)

crop_data$Cluster <- as.factor(km_final$cluster)

# Show how crops are distributed inside clusters
table(crop_data$Cluster, crop_data$Crop)
```
# 6. Visualize the clusters using PCA


```{r}
pca_fit <- prcomp(cluster_data_scaled)

pca_df <- data.frame(
  PC1 = pca_fit$x[, 1],
  PC2 = pca_fit$x[, 2],
  Cluster = crop_data$Cluster,
  Crop = crop_data$Crop
)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "K-means clusters of crop-growing conditions",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()
```


# Classification part


# 7. Create training and testing data

```{r}
set.seed(4050)

crop_split <- initial_split(crop_data, prop = 0.80, strata = Cluster)

train_data <- training(crop_split)
test_data  <- testing(crop_split)

# Use only the original predictors for classification
train_model <- train_data %>%
  dplyr::select(
    Cluster,
    Nitrogen,
    Phosphorus,
    Potassium,
    Temperature,
    Humidity,
    pH_Value,
    Rainfall
  )

test_model <- test_data %>%
  dplyr::select(
    Cluster,
    Nitrogen,
    Phosphorus,
    Potassium,
    Temperature,
    Humidity,
    pH_Value,
    Rainfall
  )
```

# 8. Set cross-validation control


```{r}




set.seed(4050)

cv_ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 3
)
```

# 9. Decision Tree with hyperparameter tuning

```{r}
tree_grid <- expand.grid(
  cp = seq(0.0005, 0.02, length.out = 20)
)

set.seed(4050)
tree_model <- train(
  Cluster ~ .,
  data = train_model,
  method = "rpart",
  trControl = cv_ctrl,
  tuneGrid = tree_grid,
  metric = "Accuracy"
)

tree_model
tree_model$bestTune

ggplot(tree_model) +
  labs(title = "Decision Tree tuning results")

rpart.plot(tree_model$finalModel)
```

# 10. Test performance for Decision Tree

```{r}
tree_pred <- predict(tree_model, newdata = test_model)

tree_cm <- confusionMatrix(
  data = tree_pred,
  reference = test_model$Cluster
)

tree_cm
```

# 11. Random Forest with hyperparameter tuning

```{r}
rf_grid <- expand.grid(
  mtry = 2:7,
  splitrule = "gini",
  min.node.size = c(1, 3, 5, 10)
)

set.seed(4050)
rf_model <- train(
  Cluster ~ .,
  data = train_model,
  method = "ranger",
  trControl = cv_ctrl,
  tuneGrid = rf_grid,
  metric = "Accuracy",
  num.trees = 500,
  importance = "impurity"
)

rf_model
rf_model$bestTune

ggplot(rf_model) +
  labs(title = "Random Forest tuning results")

varImp(rf_model)
```


# Test performance for Random Forest

```{r}
rf_pred <- predict(rf_model, newdata = test_model)

rf_cm <- confusionMatrix(
  data = rf_pred,
  reference = test_model$Cluster
)

rf_cm
```

# 13. k-NN classification with hyperparameter tuning

```{r}
knn_grid <- expand.grid(
  k = seq(3, 31, by = 2)
)

set.seed(4050)
knn_model <- train(
  Cluster ~ .,
  data = train_model,
  method = "knn",
  trControl = cv_ctrl,
  tuneGrid = knn_grid,
  metric = "Accuracy",
  preProcess = c("center", "scale")
)

knn_model
knn_model$bestTune

ggplot(knn_model) +
  labs(title = "k-NN tuning results")
```


# 14. Test performance for k-NN


```{r}
knn_pred <- predict(knn_model, newdata = test_model)

knn_cm <- confusionMatrix(
  data = knn_pred,
  reference = test_model$Cluster
)

knn_cm
```
# 15. Compare the models on the test set

```{r}
model_results <- data.frame(
  Model = c("Decision Tree", "Random Forest", "k-NN"),
  Accuracy = c(
    as.numeric(tree_cm$overall["Accuracy"]),
    as.numeric(rf_cm$overall["Accuracy"]),
    as.numeric(knn_cm$overall["Accuracy"])
  ),
  Kappa = c(
    as.numeric(tree_cm$overall["Kappa"]),
    as.numeric(rf_cm$overall["Kappa"]),
    as.numeric(knn_cm$overall["Kappa"])
  )
)

model_results <- model_results %>%
  arrange(desc(Accuracy))

print(model_results)

ggplot(model_results, aes(x = reorder(Model, Accuracy), y = Accuracy)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Test accuracy comparison of the models",
    x = "Model",
    y = "Accuracy"
  ) +
  theme_minimal()
```

# 16. Identify the best model

```{r}
best_model_name <- model_results$Model[1]
best_model_name

# Show the confusion matrix for the best model
if (best_model_name == "Decision Tree") {
  best_cm <- tree_cm
  best_pred <- tree_pred
} else if (best_model_name == "Random Forest") {
  best_cm <- rf_cm
  best_pred <- rf_pred
} else {
  best_cm <- knn_cm
  best_pred <- knn_pred
}

best_cm
```

# 17. Show a few final predictions

```{r}
final_predictions <- data.frame(
  Actual_Cluster = test_model$Cluster,
  Predicted_Cluster = best_pred
)

# Show the first 50 predictions
head(final_predictions, 50)
```



```{r}
# Or show 30 random predictions
set.seed(4050)

final_predictions %>%
  sample_n(30)
```

# 19. Create prediction data frames for each model

```{r}
library(yardstick)


tree_results <- data.frame(
  truth = test_model$Cluster,
  estimate = tree_pred
)

rf_results <- data.frame(
  truth = test_model$Cluster,
  estimate = rf_pred
)

knn_results <- data.frame(
  truth = test_model$Cluster,
  estimate = knn_pred
)

tree_results$truth <- as.factor(tree_results$truth)
tree_results$estimate <- as.factor(tree_results$estimate)

rf_results$truth <- as.factor(rf_results$truth)
rf_results$estimate <- as.factor(rf_results$estimate)

knn_results$truth <- as.factor(knn_results$truth)
knn_results$estimate <- as.factor(knn_results$estimate)
```

# 20. Overall model comparison



```{r}

tree_metrics <- dplyr::bind_rows(
  yardstick::accuracy(tree_results, truth = truth, estimate = estimate),
  yardstick::precision(tree_results, truth = truth, estimate = estimate, estimator = "macro"),
  yardstick::recall(tree_results, truth = truth, estimate = estimate, estimator = "macro"),
  yardstick::f_meas(tree_results, truth = truth, estimate = estimate, estimator = "macro")
) %>%
  dplyr::mutate(Model = "Decision Tree")

rf_metrics <- dplyr::bind_rows(
  yardstick::accuracy(rf_results, truth = truth, estimate = estimate),
  yardstick::precision(rf_results, truth = truth, estimate = estimate, estimator = "macro"),
  yardstick::recall(rf_results, truth = truth, estimate = estimate, estimator = "macro"),
  yardstick::f_meas(rf_results, truth = truth, estimate = estimate, estimator = "macro")
) %>%
  dplyr::mutate(Model = "Random Forest")

knn_metrics <- dplyr::bind_rows(
  yardstick::accuracy(knn_results, truth = truth, estimate = estimate),
  yardstick::precision(knn_results, truth = truth, estimate = estimate, estimator = "macro"),
  yardstick::recall(knn_results, truth = truth, estimate = estimate, estimator = "macro"),
  yardstick::f_meas(knn_results, truth = truth, estimate = estimate, estimator = "macro")
) %>%
  dplyr::mutate(Model = "k-NN")

all_metrics <- dplyr::bind_rows(tree_metrics, rf_metrics, knn_metrics) %>%
  dplyr::select(Model, .metric, .estimate)

print(all_metrics)

```


```{r}
library(tidyr)

metrics_table <- all_metrics %>%
  pivot_wider(names_from = .metric, values_from = .estimate)

print(metrics_table)
```



```{r}

confusionMatrix(rf_pred, test_model$Cluster)
```
# 21. Plot confusion matrix for the best model


```{r}
# Get predictions from the best model
if (best_model_name == "Decision Tree") {
  best_pred <- tree_pred
} else if (best_model_name == "Random Forest") {
  best_pred <- rf_pred
} else {
  best_pred <- knn_pred
}

best_results <- data.frame(
  Actual = test_model$Cluster,
  Predicted = best_pred
)

best_results$Actual <- factor(best_results$Actual, levels = c("1", "2", "3", "4"))
best_results$Predicted <- factor(best_results$Predicted, levels = c("1", "2", "3", "4"))

best_cm_4x4 <- table(best_results$Actual, best_results$Predicted)
best_cm_4x4

best_cm_plot <- as.data.frame(best_cm_4x4)

colnames(best_cm_plot) <- c("Actual", "Predicted", "Freq")

ggplot(best_cm_plot, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 5) +
  labs(
    title = paste("4 x 4 Confusion Matrix for Best Model:", best_model_name),
    x = "Predicted Cluster",
    y = "Actual Cluster"
  ) +
  theme_minimal()
```



```{r}

```

```{r}

```



```{r}

```




```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```


```{r}

```


```{r}

```




```{r}

```




