---
title: "Project"
author: " Keerthi, Sahaya vinish"
date: "2026-04-06"
output: html_document
---


```{r}
.libPaths()

user_lib <- path.expand("~/R/library")
if (!dir.exists(user_lib)) {
  dir.create(user_lib, recursive = TRUE)
}

.libPaths(c(user_lib, .libPaths()))

required_pkgs <- c(
  "dplyr", "ggplot2", "caret", "rsample", "rpart", "rpart.plot",
  "randomForest", "ranger", "cluster", "class", "yardstick", "tidyr"
)

missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, lib = user_lib, repos = "https://cloud.r-project.org/")
}

invisible(lapply(required_pkgs, library, character.only = TRUE))
```

```{r}

```

# 2. Read the crop dataset
```{r}
crop_data <- read.csv("Crop_Recommendation.csv")

# Check the structure
str(crop_data)
head(crop_data)

# Make sure the response variable is a factor
crop_data$Crop <- as.factor(crop_data$Crop)
```

# 3. Keep only the variables needed

```{r}
crop_data <- crop_data %>%
  dplyr::select(
    Nitrogen,
    Phosphorus,
    Potassium,
    Temperature,
    Humidity,
    pH_Value,
    Rainfall,
    Crop
  )
```

# 4. Group the observations using K-means

```{r}
# K-means uses only numeric predictors, not the crop label
cluster_data <- crop_data %>%
  dplyr::select(
    Nitrogen,
    Phosphorus,
    Potassium,
    Temperature,
    Humidity,
    pH_Value,
    Rainfall
  )

# Standardize the variables before clustering
cluster_data_scaled <- scale(cluster_data)
```

# 5. Find a good number of clusters using silhouette score


```{r}
set.seed(4050)

k_values <- 2:10
sil_scores <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  km_fit <- kmeans(cluster_data_scaled, centers = k_values[i], nstart = 25)
  sil <- silhouette(km_fit$cluster, dist(cluster_data_scaled))
  sil_scores[i] <- mean(sil[, 3])
}

sil_results <- data.frame(
  k = k_values,
  Avg_Silhouette = sil_scores
)

print(sil_results)

ggplot(sil_results, aes(x = k, y = Avg_Silhouette)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "Choosing the number of clusters using silhouette score",
    x = "Number of clusters (k)",
    y = "Average silhouette score"
  ) +
  theme_minimal()

best_k <- sil_results$k[which.max(sil_results$Avg_Silhouette)]
best_k
```


# 6. Fit the final K-means model

```{r}
set.seed(4050)
km_final <- kmeans(cluster_data_scaled, centers = best_k, nstart = 25)

crop_data$Cluster <- as.factor(km_final$cluster)

# Show how crops are distributed inside clusters
table(crop_data$Cluster, crop_data$Crop)

```



# 7. Visualize the clusters using PCA

```{r}
pca_fit <- prcomp(cluster_data_scaled)
pca_df <- data.frame(
  PC1 = pca_fit$x[, 1],
  PC2 = pca_fit$x[, 2],
  Cluster = crop_data$Cluster,
  Crop = crop_data$Crop
)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "K-means clusters of crop-growing conditions",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()
```

# Classification part

# 8. Create training and testing data

```{r}
set.seed(4050)

crop_split <- initial_split(crop_data, prop = 0.80, strata = Crop)

train_data <- training(crop_split)
test_data  <- testing(crop_split)

# Use only the original predictors for classification
train_model <- train_data %>%
  dplyr::select(
    Crop,
    Nitrogen,
    Phosphorus,
    Potassium,
    Temperature,
    Humidity,
    pH_Value,
    Rainfall
  )

test_model <- test_data %>%
  dplyr::select(
    Crop,
    Nitrogen,
    Phosphorus,
    Potassium,
    Temperature,
    Humidity,
    pH_Value,
    Rainfall
  )
```

# 9. Set cross-validation control



```{r}
set.seed(4050)

cv_ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 3
)
```



# 10. Decision Tree with hyperparameter tuning

```{r}
tree_grid <- expand.grid(
  cp = seq(0.0005, 0.02, length.out = 20)
)

set.seed(4050)
tree_model <- train(
  Crop ~ .,
  data = train_model,
  method = "rpart",
  trControl = cv_ctrl,
  tuneGrid = tree_grid,
  metric = "Accuracy"
)

tree_model
tree_model$bestTune

ggplot(tree_model) +
  labs(title = "Decision Tree tuning results")

rpart.plot(tree_model$finalModel)

```



# 11. Test performance for Decision Tree


```{r}
tree_pred <- predict(tree_model, newdata = test_model)

tree_cm <- confusionMatrix(
  data = tree_pred,
  reference = test_model$Crop
)

tree_cm
```


# 12. Random Forest with hyperparameter tuning

```{r}
rf_grid <- expand.grid(
  mtry = 2:7,
  splitrule = "gini",
  min.node.size = c(1, 3, 5, 10)
)

set.seed(4050)
rf_model <- train(
  Crop ~ .,
  data = train_model,
  method = "ranger",
  trControl = cv_ctrl,
  tuneGrid = rf_grid,
  metric = "Accuracy",
  num.trees = 500,
  importance = "impurity"
)

rf_model
rf_model$bestTune

ggplot(rf_model) +
  labs(title = "Random Forest tuning results")

varImp(rf_model)

```


# 13. Test performance for Random Forest



```{r}
rf_pred <- predict(rf_model, newdata = test_model)

rf_cm <- confusionMatrix(
  data = rf_pred,
  reference = test_model$Crop
)

rf_cm
```



# 14. k-NN classification with hyperparameter tuning


```{r}
knn_grid <- expand.grid(
  k = seq(3, 31, by = 2)
)

set.seed(4050)
knn_model <- train(
  Crop ~ .,
  data = train_model,
  method = "knn",
  trControl = cv_ctrl,
  tuneGrid = knn_grid,
  metric = "Accuracy",
  preProcess = c("center", "scale")
)

knn_model
knn_model$bestTune

ggplot(knn_model) +
  labs(title = "k-NN tuning results")

```

# 15. Test performance for k-NN

```{r}
knn_pred <- predict(knn_model, newdata = test_model)

knn_cm <- confusionMatrix(
  data = knn_pred,
  reference = test_model$Crop
)

knn_cm
```


# 16. Compare the models on the test set


```{r}
model_results <- data.frame(
  Model = c("Decision Tree", "Random Forest", "k-NN"),
  Accuracy = c(
    as.numeric(tree_cm$overall["Accuracy"]),
    as.numeric(rf_cm$overall["Accuracy"]),
    as.numeric(knn_cm$overall["Accuracy"])
  ),
  Kappa = c(
    as.numeric(tree_cm$overall["Kappa"]),
    as.numeric(rf_cm$overall["Kappa"]),
    as.numeric(knn_cm$overall["Kappa"])
  )
)

model_results <- model_results %>%
  arrange(desc(Accuracy))

print(model_results)

ggplot(model_results, aes(x = reorder(Model, Accuracy), y = Accuracy)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Test accuracy comparison of the models",
    x = "Model",
    y = "Accuracy"
  ) +
  theme_minimal()
```

# 17. Identify the best model


```{r}
best_model_name <- model_results$Model[1]
best_model_name

# Show the confusion matrix for the best model
if (best_model_name == "Decision Tree") {
  best_cm <- tree_cm
  best_pred <- tree_pred
} else if (best_model_name == "Random Forest") {
  best_cm <- rf_cm
  best_pred <- rf_pred
} else {
  best_cm <- knn_cm
  best_pred <- knn_pred
}

best_cm

```


# 18. Show a few final predictions


```{r}
final_predictions <- data.frame(
  Actual_Crop = test_model$Crop,
  Predicted_Crop = best_pred
)

# Show the first 50 predictions
head(final_predictions, 50)


```



```{r}
# Or show 30 random predictions
set.seed(4050)
final_predictions %>%
  sample_n(30)
```



```{r}


library(yardstick)
library(dplyr)
```

# 20. Create prediction data frames for each model


```{r}
tree_results <- data.frame(
  truth = test_model$Crop,
  estimate = tree_pred
)

rf_results <- data.frame(
  truth = test_model$Crop,
  estimate = rf_pred
)

knn_results <- data.frame(
  truth = test_model$Crop,
  estimate = knn_pred
)

# Make sure both are factors
tree_results$truth <- as.factor(tree_results$truth)
tree_results$estimate <- as.factor(tree_results$estimate)

rf_results$truth <- as.factor(rf_results$truth)
rf_results$estimate <- as.factor(rf_results$estimate)

knn_results$truth <- as.factor(knn_results$truth)
knn_results$estimate <- as.factor(knn_results$estimate)
```


# 21. Overall model comparison

```{r}
tree_metrics <- bind_rows(
  accuracy(tree_results, truth = truth, estimate = estimate),
  precision(tree_results, truth = truth, estimate = estimate, estimator = "macro"),
  recall(tree_results, truth = truth, estimate = estimate, estimator = "macro"),
  f_meas(tree_results, truth = truth, estimate = estimate, estimator = "macro")
) %>%
  mutate(Model = "Decision Tree")

rf_metrics <- bind_rows(
  accuracy(rf_results, truth = truth, estimate = estimate),
  precision(rf_results, truth = truth, estimate = estimate, estimator = "macro"),
  recall(rf_results, truth = truth, estimate = estimate, estimator = "macro"),
  f_meas(rf_results, truth = truth, estimate = estimate, estimator = "macro")
) %>%
  mutate(Model = "Random Forest")

knn_metrics <- bind_rows(
  accuracy(knn_results, truth = truth, estimate = estimate),
  precision(knn_results, truth = truth, estimate = estimate, estimator = "macro"),
  recall(knn_results, truth = truth, estimate = estimate, estimator = "macro"),
  f_meas(knn_results, truth = truth, estimate = estimate, estimator = "macro")
) %>%
  mutate(Model = "k-NN")

all_metrics <- bind_rows(tree_metrics, rf_metrics, knn_metrics) %>%
  select(Model, .metric, .estimate)

print(all_metrics)
```




```{r}
library(tidyr)

metrics_table <- all_metrics %>%
  pivot_wider(names_from = .metric, values_from = .estimate)

print(metrics_table)
```


```{r}
# Example: Random Forest confusion matrix
confusionMatrix(rf_pred, test_model$Crop)
```

# Plot confusion matrix for the best model

```{r}
# Get predictions from the best model
if (best_model_name == "Decision Tree") {
  best_pred <- tree_pred
} else if (best_model_name == "Random Forest") {
  best_pred <- rf_pred
} else {
  best_pred <- knn_pred
}

# Create confusion matrix table
cm_plot_data <- data.frame(
  Actual = test_model$Crop,
  Predicted = best_pred
) %>%
  count(Actual, Predicted)

# Plot confusion matrix
ggplot(cm_plot_data, aes(x = Predicted, y = Actual, fill = n)) +
  geom_tile() +
  geom_text(aes(label = n), size = 3) +
  labs(
    title = paste("Confusion Matrix -", best_model_name),
    x = "Predicted Crop",
    y = "Actual Crop",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```




```{r}
# ============================================================
# 20. Better-looking confusion matrix plot
# ============================================================
cm_plot_data <- data.frame(
  Actual = test_model$Crop,
  Predicted = best_pred
) %>%
  count(Actual, Predicted)

ggplot(cm_plot_data, aes(x = Predicted, y = Actual, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = paste("Confusion Matrix for Best Model:", best_model_name),
    x = "Predicted Class",
    y = "Actual Class",
    fill = "Frequency"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

